---
title: "Income Housing Affordability"
author: "Hien Vo & Hai Nguyen"
date: "2024-04-26"
output: html_document
---
hh
Research topic: Income Housing Affordability in California
The main question: Is Californian’s income sufficient to afford to buy a house?

Sub-questions:

1/ How has the median house price changed in California since 2018? Which county
has experienced the highest increase in house pricing over the 2018-now period?
2/ How has the median personal income changed in California since 2018? In which
county has the highest median personal income now?
3/ What is the ratio of median house price to median personal income in each
county? In which county Californians can be easier to buy a house based on their
salary?
4/ What is the growth in house prices compared to median personal incomes?
5/ How has the mortgage rate changed since 2018? Create a line plot and interpret
the relaƟonship between the median house price and mortgage

Variable definition

SHF           :   Single Family Home Raw Data
Condo         :   Condo Raw Data
SFH_Price     :   Single Family Home Price Data in the US from 2018
SFH_CA        :   Single Family Home Price Data in California from 2018
Condo_Price   :   Condo Price Data in the US from 2018
Condo_CA      :   Condo Price Data in California from 2018
Join_HP_CA    :   Merge Data House Prices (including SFH and Condo) in California from 2018
Median_HP_CA  :   Median House Price (Including SFH and Condo) in California from 2018


```{r}
library(dplyr, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(ggplot2, quietly = TRUE)


# read the Single Family House Data
SFH <- read.csv("D:/SP__24/MATH 167R/Project/167R/Single Family House.csv")
# read the Single Condo Data
Condo <- read.csv("D:/SP__24/MATH 167R/Project/167R/Condo.csv")

```

A. Filter House Prices Data

1. Filter Single Family House Data
```{r}
# Find the index of the column of Jan 2018
index_Jan2018 <- grep("2018", colnames(SFH))[1] #[1] to get the first col in 2018
    
# Extract the data since Jan 2018 to the latest
Temp_SFH <- SFH|>
 select(3, 6, 7, 8, all_of(index_Jan2018_Condo) : ncol(SFH)) 
             
# Convert the table from wide to long format as the years are shown in columns
SFH_Price <- pivot_longer(
                          Temp_SFH,
                          cols = 5 : ncol(Temp_SFH),
                          names_to = "Date",
                          values_to = "Price")

# Format date and change column's names
SFH_Price <- SFH_Price |>
             mutate(Date = as.Date(strptime(Date, format = "X%m.%d.%Y")))

names(SFH_Price) <- c("City", "State", "Metro", "County", "Date", "SFH_Price")

# Note: SFH_Price is the table including all single family homes prices of the US by City from 2018 to present.

SFH_CA <- SFH_Price |>  # single family homes prices in California from 2018 to present.
          filter(State == "CA", na.rm = TRUE)

```


2. Filter Single Condo Data
```{r}
# Find the index of the column of Jan 2018
index_Jan2018_Condo <- grep(2018, colnames(Condo))[1] #[1] to get the first col in 2018
    
# Extract the data since Jan 2018 to the latest
Temp_Condo <- Condo|>
              select(3, 6, 7, 8, all_of(index_Jan2018_Condo) : ncol(Condo)) 
              #col 3 = City name/Region Name, #col 6 = State Name, $col 7 = Metro; #col 8 = County

# Convert the table from wide to long format as the years are shown in columns
Condo_Price <- pivot_longer(
                          Temp_Condo,
                          cols = 5 : ncol(Temp_Condo),
                          names_to = "Date",
                          values_to = "Price")
# Format date and change column's names
Condo_Price <- Condo_Price |>
               mutate(Date = as.Date(strptime(Date, format = "X%m.%d.%Y")))

names(Condo_Price) <- c("City", "State", "Metro", "County", "Date", "Condo_Price")

# Note: SFH_Price is the table including all Condo prices of the US by City from 2018 to present.

Condo_CA <- Condo_Price |>  # Condo prices in California from 2018 to present.
            filter(State == "CA", na.rm = TRUE)

```

3. Merge data: SFH and Condo Prices in California

```{r}
Join_HP_CA <- merge(SFH_CA, Condo_CA, by = c("City", "State", "Metro", "County", "Date"), all = TRUE)

```

B. 
1/ How has the median house price changed in California since 2018?

```{r}
# Find Median Price
Median_HP_CA <- Join_HP_CA |>
                 mutate(Year = format(Date, "%Y")) |>
                 group_by(Year) |>
                 summarize (Median_SFH_Price = median(SFH_Price, na.rm = TRUE),
                            Median_Condo_Price = median(Condo_Price, na.rm = TRUE))

Median_HP_CA$Year <- as.numeric(Median_HP_CA$Year) # convert Year Column to numeric

ggplot(Median_HP_CA, aes(x = Year)) +
  geom_line(aes(y = Median_SFH_Price, color = "Median_SFH_Price")) +
  geom_line(aes(y = Median_Condo_Price, color = "Median_Condo_Price")) +
  labs(title = "House Prices in California from 2018 to Present",
       x = "Year",
       y = "Median Price") +
  scale_color_manual(values = c("Median_SFH_Price" = "blue", "Median_Condo_Price" = "red")) +
  theme_minimal()

```


