---
title: "Income Housing Affordability"
author: "Hien Vo & Hai Nguyen"
date: "2024-04-26"
output: html_document
---

test1

Research topic: Income Housing Affordability in California
The main question: Is Californian’s income sufficient to afford to buy a house?

Sub-questions:

1/ How has the median house price changed in California since 2018? Which county
has experienced the highest increase in house pricing over the 2018-now period?
2/ How has the median personal income changed in California since 2018? In which
county has the highest median personal income now?
3/ What is the ratio of median house price to median personal income in each
county? In which county Californians can be easier to buy a house based on their
salary?
4/ What is the growth in house prices compared to median personal incomes?
5/ How has the mortgage rate changed since 2018? Create a line plot and interpret
the relaƟonship between the median house price and mortgage

```{r, message=FALSE}
library(dplyr, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(tidyselect, quietly = TRUE)

# read the Single Family House Data
MORTGAGE15US <- readr::read_csv("MORTGAGE15US.csv")
MORTGAGE30US <- readr::read_csv("MORTGAGE30US.csv")
Condo <- readr::read_csv("Condo.csv")
SFH <- readr::read_csv("Single Family House.csv")

```

A. Filter House Prices Data

1. Filter Single Family House Data
```{r}
# Find the index of the column of Jan 2018
index_Jan2018 <- grep(2018, colnames(SFH))[1] #[1] to get the first col in 2018
    
# Extract the data since Jan 2018 to the latest
Temp_SFH <- SFH|>
                select(3, 5, 7, 8, index_Jan2018:ncol(SFH)) 
              #col 3 = City name/Region Name, #col 5 = State Name, $col 7 = Metro; #col 8 = County

# Convert the table from wide to long format as the years are shown in columns
SFH_Price <- pivot_longer(
                          Temp_SFH,
                          cols = 5 : ncol(Temp_SFH),
                          names_to = "Date",
                          values_to = "Price")
# Format date
SFH_Price <- SFH_Price |>
                mutate(Date = as.Date(strptime(Date, format = "X%m.%d.%Y")))

names(SFH_Price) <- c("City", "State", "Metro", "County", "Date", "SFH Price")

# Note: SFH_Price is the table including all single family homes prices of the US by City from 2018 to present.

SFH_CA <- SFH_Price |>  # single family homes prices in California from 2018 to present.
                  filter(State == "CA", na.rm = TRUE)

```

```{r}
library(tidyverse)
library(dplyr)
 
```




2. Filter Single Condo Data
```{r}
# Find the index of the column of Jan 2018
index_Jan2018_Condo <- grep(2018, colnames(Condo))[1] #[1] to get the first col in 2018
    
# Extract the data since Jan 2018 to the latest
Temp_Condo <- Condo|>
                select(3, 5, 7, 8, index_Jan2018_Condo:ncol(Condo)) 
              #col 3 = City name/Region Name, #col 5 = State Name, $col 7 = Metro; #col 8 = County

# Convert the table from wide to long format as the years are shown in columns
Condo_Price <- pivot_longer(
                          Temp_Condo,
                          cols = 5 : ncol(Temp_Condo),
                          names_to = "Date",
                          values_to = "Price")
# Format date
Condo_Price <- Condo_Price |>
                mutate(Date = as.Date(strptime(Date, format = "X%m.%d.%Y")))
names(Condo_Price) <- c("City", "State", "Metro", "County", "Date", "Condo Price")

# Note: SFH_Price is the table including all Condo prices of the US by City from 2018 to present.

Condo_CA <- Condo_Price |>  # Condo prices in California from 2018 to present.
                  filter(State == "CA", na.rm = TRUE)

Condo_mean_price <- mean(Condo_CA$`Condo Price`)

```

3. Merge data

```{r}
Join_HP <- merge(SFH_Price, Condo_Price, by = c("City"), na.rm = TRUE, all = TRUE)
```

B. 
1/ How has the median house price changed in California since 2018?

```{r}
Median_SFH_CA <- SFH_CA |>
                 mutate(Year = format(Date, "%Y")) |>
                 group_by(Year) |>
                 summarize (Median_SFH_CA = median(Price, na.rm = TRUE))

Median_Condo_CA <- Condo_CA |>
                 mutate(Year = format(Date, "%Y")) |>
                 group_by(Year) |>
                 summarize (Median_Condo_CA = median(Price, na.rm = TRUE))


```


